---
title: 'Class 17: COVID-19 Vaccination Rates'
author: "Ethan Harding (PID A15468670)"
date: "11/23/2021"
output:
  html_document: default
  pdf_document: default
---

# Getting Started

First, import and read the vaccination data. 
```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> Q1. What column details the total number of people fully vaccinated?

-- persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area?

-- zip_code_tabulation_area

> Q3. What is the earliest date in this dataset?

```{r}
head(vax$as_of_date, 1)
```

> Q4. What is the latest date in this dataset?

```{r}
tail(vax$as_of_date, 1)
```

Let's call the `skim()` function from the `skimr` package to get a quick overview of the dataset. 
```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset?

-- 9

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

```{r}
sum( is.na(vax$persons_fully_vaccinated) )
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
round( sum( is.na(vax$persons_fully_vaccinated) ) / nrow(vax) * 100, 2 )

```

> Q8. [Optional]: Why might this data be missing?

-- Some zip codes include areas that with federal agencies, whose data is not included in the CDC's vaccination rate file. 


We will use the **lubridate** package to make life a lot easier when dealing with dates and times. 

```{r}
library(lubridate)
```

Here, we make our `as_of_date` column lubridate format.
```{r}
# Specify that we are using the Year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now I can do useful math with dates more easily: 

```{r}
today() - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[ nrow(vax) ]
```

> Q. How many days between the first and the last entry in the dataset? 

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```


> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length( unique(vax$as_of_date))
```

This sounds good
```{r}
46*7
```

# Working with Zip Codes

In R we can use the zipcodeR package to make working with these codes easier.
```{r}
library(zipcodeR)
```

```{r}
reverse_zipcode(c('92037', "92109") )
```

# Focus on the San Diego County
```{r}
table(vax$county)
```

We will subset with base R. 
```{r}
sd <- vax$county == "San Diego"
head(vax[sd,])
```

But let's use the **dplyr** package and it's **filter()** function: 
```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")
head(sd)
```

> Q. How many entries are there for San Diego county? 

```{r}
nrow(sd)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length( unique(sd$zip_code_tabulation_area) )
```

Using dplyr is often more convenient when we are subsetting across multiple criteria - for example all San Diego county areas with a population of over 10,000.

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
head(sd.10)
```


> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]
```

> What is the population in the 92037 ZIP code area?

```{r}
filter(sd, zip_code_tabulation_area == "92037")[1,]
```

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”?

```{r}
sd.now <- filter(sd, as_of_date == "2021-11-09")

mean(sd.now$percent_of_population_fully_vaccinated,
     na.rm=TRUE)
```

We can look at the 6-number summary:
```{r}
summary( sd.now$percent_of_population_fully_vaccinated)
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

Using base R plots:
```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

Using ggplot: 
```{r}
library(ggplot2)

ggplot(sd.now) + 
  aes(percent_of_population_fully_vaccinated) + 
  geom_histogram(bins=15)
```

What about 92037 - UCSD / La Jolla?

```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")
head(ucsd)
```


# Time series of vaccination rate for 92037 

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", title="Vaccination rates for La Jolla 92037")
```

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2021-11-16”.

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)
```

> How many unique zip codes have a population as large as 92037?

```{r}
length(unique(vax.36$percent_of_population_fully_vaccinated))
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", title="Vaccination rates for La Jolla 92037") +
  geom_hline(yintercept = 0.6629812, color="red", linetype=2)
 
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  labs(x="Percent Vaccinated", y="Count")
```

>Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

92040 is below the average and 92109 is above average. 
```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```



> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rates of California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = 0.66, linetype=2, col="red")
```

> Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

I'm fortunate enough that all of my family is local, so I will not be traveling far for Thanksgiving, and as long as people are vaccinated and getting frequent COVID tests if they are traveling far / meeting family that are, I am fine with in-person class next week. Obviously, I would wish those that are showing symptoms to stay home. 





