---
title: "Halloween Mini Project"
author: "Ethan Harding (PID A15468670)"
date: "10/28/2021"
output:
  pdf_document: default
  html_document: default
---

## 1. Importing Candy Data
```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

> **Q1.** How many different candy types are in this dataset?

```{r}
dim(candy)
```
There are 85 different candy types in this dataset. 

> **Q2.** How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in the dataset. 

## 2. What is your favorite candy?

> **Q3.** What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["ReeseÕs Peanut Butter cup", ]$winpercent
```
My favorite candy is in this dataset is Reese's Peanut Butter cups and its winpercent value is 84.18029%. 

> **Q4.** What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
The winpercent value for Kit Kat is 76.7686%.

> **Q5.** What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
The winpercent vallue for Tootsie Roll Snack Bars is 49.6535%. 

Use skimr package give quick overview of candy dataset. 

```{r}
# install.packages("skimr") if you do not have it installed already
library("skimr")
skim(candy)
```

> **Q6.** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
 
The winpercent column looks to be on a different scale compared to the majority of the other columns in the dataset, which are on a scale from 0 to 1, whereas the winpercent column seems to be on a scale from 0 to 100. 

> **Q7.** What do you think a zero and one represent for the candy$chocolate column?

A zero represents that the specified candy does NOT have chocolate, whereas a 1 represents the candy having chocolate. 

> **Q8.** Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) + geom_histogram(bins=20)

hist(candy$winpercent)
```


> **Q9.** Is the distribution of winpercent values symmetrical?

No, the distribution of winpercent values is skewed a little to the right. 

> **Q10.** Is the center of the distribution above or below 50%?

The center of the distribution is below 50%.

> **Q11.** On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

On average, chocolate candy is ranked higher than fruit candy. 

> **Q12.** Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```
Yes, the difference is statistically significant becuase the p-value is <<0.05 (p=2.871e-08). 
 
## 3. Overall Candy Rankings
 
```{r}
head(candy[order(candy$winpercent),], n=5)
```
 
> **Q13.** What are the five least liked candy types in this set?
 
The five least liked candy types in this set are: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble and Jawbusters.
 
```{r}
head(candy[order(candy$winpercent, decreasing=TRUE),], n=5)
```


> **Q14.** What are the top 5 all time favorite candy types out of this set?
 
The top 5 all time favorite candy types out of this set are: Reese's Peanut Butter Cups, Reese's Miniatures, Twix, Kit Kat, and Snickers. 

> **Q15.** Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> **Q16.** This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

Let's add some color now. First, let's assign colors to each candy type. 
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

Use ggplot to graph with the color type. 
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> **Q17.** What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets. 

> **Q18.** What is the best ranked fruity candy?

The best ranked fruity candy is Starburst. 

## 4. Taking a look at pricepoint

Use ggrepel package to plot names next to candy datapoints. Plot price vs. winpercent. By using ggprepel instead of regular ggplot, the candy names will not overlap and instead "repel" away from each other. 
```{r}
library(ggrepel)

# Plot price vs. win percent
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3, max.overlaps = 6)
```

> **Q19.** Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The candy that offers the "most bang for your buck" in terms of being the highest ranked winpercent for the least money is  Reese's Miniatures. 

> **Q20.** What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The top 5 most expensive candies, from most to least expensive, are: Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular is also the most expensive, Nik L Nip. 

> **Q21.** Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Make a lollipop chart of the pricepercent values
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_point() + geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40")
```

## 5. Exploring the Correlation Structure

We will now use the "corrplot" package to plot a correlation matrix. 
```{r}
library("corrplot")
cij <- cor(candy)
corrplot(cij)
```

> **Q22.** Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Two variables that are anti-correlated are chocolate and fruity. 

> **Q23.** Similarly, what two variables are most positively correlated?

The two variables that are most positively correlated are chocolate and winpercent. 

## 6. Principal Component Analysis

Let's apply PCA to our candy dataset. 
```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```

Now we can plot PC1 vs PC2. 
```{r}
plot(pca$x[,1:2])
```

Let's add color. 
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

We can make a nicer plot with ggplot, but we have to create a data frame with our PCA results so that ggplot can read and plot it.  
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

# Plot using ggplot
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

Now add labels using ggrepel. 
```{r}
library(ggrepel)

p + geom_text_repel(size=3, col=my_cols, max.overlaps = 8)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (pink), other (black)",
       caption="Data from 538")
```

Use "plotly" package to generate an interactive plot. 
```{r}
library(plotly)
ggplotly(p)
```

Let's look at the PCA loadings. 
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> **Q24.** What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, Hard, and Pluribus are picked up strongly by PC1 in the positive direction. These make sense becuase fruity candy is generally hard candies that come in a bag or box with a big amount. 

